### 2020-09-10 ART_Tambor_Szydlowski/Tambor_Szydlowski-PU.tex
Final version before printing and publishing online
Minor changes in toc hyperreferences

\begin{artplenv2auth}{Marek Szydłowski, Paweł Tambor}
	{Czy fizyczne teorie efektywne są wiarygodną strategią osiągnięcia teorii ostatecznej?}
	{Czy fizyczne teorie efektywne są wiarygodną strategią\ldots}
	{Czy fizyczne teorie efektywne są wiarygodną strategią osiągnięcia teorii ostatecznej?}
	{Wydział Filozofii KUL, Uniwersytet Jagielloński Kraków}
	{Are Physical Effective Theories the Reliable Strategy for Achieving Certain Knowledge?}
	{The paper presents the methodology of effective theories as a~strategy used in the process of development of modern physics to reach a~final theory. We present the definition and characteristic features of an effective theory, as well as the answer to the question of whether and what kind of scenario of reaching a~final theory is realized by contemporary physics. We argue that the process of development of physics in the direction of a~final theory is potentially final, i.e. expressible in the conceptual schema of effective theories and as such it is convergent to a~final theory. In each effective theory there are physical constants, however, whose status differs from logical constants. They have a~dimension (length, energy, etc.) and are used to compare physical quantities. The structure of relevant effective theory can be interpreted in the epistemological framework of approximated truth theory. In the case study of cosmological models the sequence of models is convergent to potentially true model. The Standard Cosmological Model is the theory of the structure and dynamics of the Universe.}
	{philosophy and methodology of science, effective theory, final theory, cosmology, approximated truth theories.}
	{%
			{\flushright\subbold{Marek Szydłowski}\\\subsubsectit\small{Uniwersytet Jagielloński}\par}%
			{\flushright\subbold{Paweł Tambor}\\\subsubsectit\small{Katolicki Uniwersytet Lubelski}\par}%
		}

@@

Szynkiewicz, odnosząc się~do Poppera, określa teorię~ostateczną~jako: ,,[...] propozycję~stanowiącą końcowy, pełny i~niemodyfikowalny fragment wiedzy teoretycznej. Koncepcje tego typu, z~uwagi na ich zakres, dzielone są zazwyczaj na dwie zasadnicze klasy: 1) teorie wszystkiego -- najbardziej ambitne i~odnoszące się do wszystkich klas zjawisk, 2) teorie ostateczne (cząstkowe) -- opisujące wybrane, ograniczone obszary przedmiotowe (np. zjawiska fizyczne, biologiczne). [...] Teoria ostateczna byłaby więc w~proponowanym ujęciu koncepcją~posiadającą wewnętrznie niesprzeczną~i niemodyfikowalną aksjomatykę.''
%\label{ref:RND4RBAmt5LFF}(Szynkiewicz, 2011, s.~1274–1275).
\parencite[][s.~1274–1275]{szynkiewicz_teorie_2011}. %
 Autor dookreśla definicję poprzez podanie istotnych cech teorii ostatecznej: zupełność (teoria odnosi się~do wszystkich zjawisk w~opisywanej przez siebie domenie), prostota logiczna (wypadkowa informacyjnej zawartości teorii i~ilości postulatów wyjściowych), stabilność i~niezmienność~strukturalna teorii.

>>>>

\begin{artplenv2auth}{Marek Szydłowski, Paweł Tambor}
	{Czy fizyczne teorie efektywne są wiarygodną strategią osiągnięcia teorii ostatecznej?}
	{Czy fizyczne teorie efektywne są wiarygodną strategią\ldots}
	{Czy fizyczne teorie efektywne są wiarygodną strategią osiągnięcia teorii ostatecznej?}
	{Wydział Filozofii KUL, Uniwersytet Jagielloński Kraków}
	{Are physical effective theories the reliable strategy for achieving certain knowledge?}
	{The paper presents the methodology of effective theories as a~strategy used in the process of development of modern physics to reach a~final theory. We present the definition and characteristic features of an effective theory, as well as the answer to the question of whether and what kind of scenario of reaching a~final theory is realized by contemporary physics. We argue that the process of development of physics in the direction of a~final theory is potentially final, i.e. expressible in the conceptual schema of effective theories and as such it is convergent to a~final theory. In each effective theory there are physical constants, however, whose status differs from logical constants. They have a~dimension (length, energy, etc.) and are used to compare physical quantities. The structure of relevant effective theory can be interpreted in the epistemological framework of approximated truth theory. In the case study of cosmological models the sequence of models is convergent to potentially true model. The Standard Cosmological Model is the theory of the structure and dynamics of the Universe.}
	{philosophy and methodology of science, effective theory, final theory, cosmology, approximated truth theories.}
	{%
			{\flushright\subbold{Marek Szydłowski}\\\subsubsectit\small{Uniwersytet Jagielloński}\par}%
			{\flushright\subbold{Paweł Tambor}\\\subsubsectit\small{Katolicki Uniwersytet Lubelski}\par}%
		}

@@

Szynkiewicz, odnosząc się~do Poppera, określa teorię~ostateczną~jako: ,,[...] propozycję~stanowiącą końcowy, pełny i~niemodyfikowalny fragment wiedzy teoretycznej. Koncepcje tego typu, z~uwagi na ich zakres, dzielone są zazwyczaj na dwie zasadnicze klasy: 1) teorie wszystkiego -- najbardziej ambitne i~odnoszące się do wszystkich klas zjawisk, 2) teorie ostateczne (cząstkowe) -- opisujące wybrane, ograniczone obszary przedmiotowe (np. zjawiska fizyczne, biologiczne). [...] Teoria ostateczna byłaby więc w~proponowanym ujęciu koncepcją~posiadającą wewnętrznie niesprzeczną~i niemodyfikowalną aksjomatykę''
%\label{ref:RND4RBAmt5LFF}(Szynkiewicz, 2011, s.~1274–1275).
\parencite[][s.~1274–1275]{szynkiewicz_teorie_2011}. %
 Autor dookreśla definicję poprzez podanie istotnych cech teorii ostatecznej: zupełność (teoria odnosi się~do wszystkich zjawisk w~opisywanej przez siebie domenie), prostota logiczna (wypadkowa informacyjnej zawartości teorii i~ilości postulatów wyjściowych), stabilność i~niezmienność~strukturalna teorii.
